---
title: Which country spends best on healthcare?
author: Osvaldo
date: '2018-01-21'
headerimg: "/img/surfers_light_blur.jpg"
slug: which-country-spends-best-on-healthcare
categories:
  - data
tags:
  - OECD
  - country statistics
  - healthcare
---

In its publication [Health at a Glance 2017](http://dx.doi.org/10.1787/health_glance-2017-en), the most famous
rich countries think-tank (OECD) presents a trove of data
on health status and health system performance from many countries. Here below I report a small analysis I did looking at
life expectancy and health spending per capita. The analysis below is performed on data available [here](http://www.oecd-ilibrary.org/social-issues-migration-health/health-at-a-glance-2017/life-expectancy-at-birth-and-health-spending-per-capita-2015-or-nearest-year_health_glance-2017-graph8-en), including the 35 OECD members plus some non-members like India, Indonesia, China, Costa Rica, Russia and others.

```{r all_observed, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(plotly)
library(widgetframe)
life_exp_spending <- read_csv("life_exp_spending.csv") %>%
  arrange(health_spending_USD_ppp)
# kable(life_exp_spending, caption = 'Health spending and life expectancy at birth, 2015 (or nearest year)')

p <- plot_ly(life_exp_spending, x = ~health_spending_USD_ppp, y = ~life_expectancy, type = 'scatter',
        mode = 'markers', text = ~country) %>%
        # textposition = 'middle right', textfont = list(color = '#000000', size = 16)) %>%
  layout(title = 'Health spending per capita and life expectancy at birth, 2015 (or nearest year)',
         xaxis = list(title = 'Health spending per capita (USD at PPP)',
                      zeroline = FALSE),
         yaxis = list(title = 'Life expectancy at birth (years)'))

frameWidget(p)
```

USA are known to spend the most amont the OECD countries. Nevertheless, despite spending 9507 USD per capita, their life
expectancy is not in the top-tier and achieves 78.8 years. Further, it has been going down since a couple of years as
a conseguence of the [opioid crisis](https://www.theguardian.com/us-news/2017/dec/21/us-life-expectancy-down-for-second-year-in-a-row-amid-opioid-crisis). Another outlier is South Africa, whose spending is in the low range of the pack, but achieves
the lowest life expectancy in the countries considered (among other factors, HIV prevalence among adults in
South Africa is around 19%).

Except for these notable outliers, there seems to be a clear correlation between spending and life expectancy, with the latter
increasing until 3000-4000 dollars and then flattening at around 83 years. Something that could be captured with a reasonable
goodness-of-fit by an exponential response as in the next plot (USA and South Africa have been removed).

```{r predicted, echo=FALSE, message=FALSE, warning=FALSE}
life_exp_spending <- life_exp_spending %>%
  filter(!(country_code %in% c('USA', 'ZAF')))

fit <- nls(life_expectancy ~ le_max * (1 - exp(a - b * health_spending_USD_ppp)),
           data = life_exp_spending,
           start = list(le_max = 80, a = 10, b = 0.001))
# summary(fit)
# plot(life_exp_spending$health_spending_USD_ppp, life_exp_spending$life_expectancy)
# lines(life_exp_spending$health_spending_USD_ppp, predict(fit))
life_exp_spending$fit <- predict(fit)
life_exp_spending$residuals <- round(residuals(fit), 1)

p <- plot_ly(life_exp_spending, x = ~health_spending_USD_ppp, y = ~life_expectancy, type = 'scatter',
             mode = 'markers', text = ~country, name = 'observed') %>%
  add_trace(y = ~fit, name = 'predicted', mode = 'lines') %>%
  layout(title = 'Health spending per capita and life expectancy at birth, 2015 (or nearest year)',
         xaxis = list(title = 'Health spending per capita (USD at PPP)',
                      zeroline = FALSE),
         yaxis = list(title = 'Life expectancy at birth (years)'))


# life_exp_spending <- life_exp_spending %>%
#   arrange(desc(residuals))
frameWidget(p)
```

This brings us to the final plot, which shows the over- and under-performers: which country, given its spending, achieves
life expectancy above and below the _predicted_ one?

```{r residuals, echo=FALSE, message=FALSE, warning=FALSE}

life_exp_spending <- life_exp_spending %>%
  arrange(desc(residuals))
life_exp_spending$country_code <- factor(life_exp_spending$country_code, levels = life_exp_spending$country_code)
p <- plot_ly(life_exp_spending, x = ~country_code, y = ~residuals, type="bar", text=life_exp_spending$country) %>%
    layout(title = 'Residual (observed - predicted) life expectancy',
         xaxis = list(title=''),
         yaxis = list(title = 'Excess life expectancy (years)'))

frameWidget(p)
```

In this special ranking, the worst performers are Russia, Lithuania, Hungary, Slovakia, and Latvia; with spendings between 1351
and 2059 USD, they fare between 2 and five years worse than predicted. Interestingly, they are trailed by Germany, Denmark,
and Belgium; with spendings between 4778 and  5353 USD, they are 1-1.5 years below predictions.
Turkey, China, and Costa Rica, although occupying the low-spending positions and, at the same time, not featuring the highest
life expectancy, do much better than the prediction and are the winners in this special ranking. A group of countries with the
highest life expectancy and with moderate spending is represented by Japan (the highest life expectancy overall), Spain, Italy, Israel, and Korea.

### Conclusions

A few things one should keep in mind in order not to take this analysis too seriously:

- it's just a fit, an attempt to overlay a very simple curve over a humongously complex system like healthcare,
- although life expectancy is an important measure of the overall quality of a country healthcare, other measures exist
  that are also important to the patient,
- PPP (purchase power parity) tries to capture the different level of costs in two different countries by comparing the cost
  of a basket of goods, but it clearly has limitations,
- it's just a fit.


Header photo by [Barry Thomas](https://flic.kr/p/e8j6WN) with modifications.